---
title: "DATA363 Project"
output: html_document
date: "2023-03-12"
---

Import data files
```{r}
install.packages('dplyr')
library(dplyr)

AveIncome <- read.csv("DATA363 Project - Income.csv", header = TRUE)
#Imports Average Income Data set

AveTuition <-read.csv("DATA363 Project - College.csv", header = TRUE)
#Imports Average State Tuition Data set
```

Filter out unneeded college data
```{r}
AveTuition2018 <- filter(AveTuition, Year %in% c("2018") & Length %in% c("4-year"))
#Filters Average Tuition for the year 2018 and only 4-year long programs
```

Filter out unneeded income data
```{r}
AveIncomeState <- AveIncome[4, ] 
#Splits of row 4 containing column names for state and year

AveIncomeState2 <- AveIncome[7:57, ]
 #Splits dataset into 50 states and their average incomes

AveIncomeStateTrue <- rbind(AveIncomeState, AveIncomeState2)
#combines row 4 with row 7:57

NewAveIncome <- AveIncomeStateTrue[2:52, ] 
#Creates a new dataset

colnames(NewAveIncome) <- AveIncomeStateTrue[1, ] 
#Titles new dataset with row 4 of AveIncome

AveIncome2018 <- subset(NewAveIncome, select = c("State", "2018")) #Creates a dataset of average state income for the year 2018

rownames(AveIncome2018) = seq(length=nrow(AveIncome2018))
#Renumbers rows properly

AveIncome2018NoWyoming <- AveIncome2018[-51,]
#Removes Wyoming to compare against private college costs

AveIncome2018NoDC <- AveIncome2018[-9,]
#Removes DC to compare against out-of-state college costs

rownames(AveIncome2018NoDC) = seq(length=nrow(AveIncome2018NoDC))
#Renumbers rows properly

colnames(AveIncome2018NoWyoming)[2] ="Income"
#Renames column "2018" to "Income"
colnames(AveIncome2018)[2] ="Income"
#Renames column "2018" to "Income"
colnames(AveIncome2018NoDC)[2] ="Income"
#Renames column "2018" to "Income"

AveIncome2018NoWyoming$Income <- as.numeric(gsub(",","", AveIncome2018NoWyoming$Income))
#Removes comma from number and turns it into double type
AveIncome2018$Income <- as.numeric(gsub(",","", AveIncome2018$Income))
#Removes comma from number and turns it into double type
AveIncome2018NoDC$Income <- as.numeric(gsub(",","", AveIncome2018NoDC$Income))
#Removes comma from number and turns it into double type

AveIncome2018NoWyoming
AveIncome2018
AveIncome2018NoDC
```

Create new college data table with fees/tuition and room/board combined
```{r}
AveCost2018 <- data.frame(matrix(ncol = 3, nrow = 0))
#Creates new empty data frame with three columns
colnames(AveCost2018) <- c('State', 'Type', 'Cost')
#Names the columns
for(i in 1:nrow(AveTuition2018)) {
  if (i %% 2 == 1) {
    AveCost2018[nrow(AveCost2018) + 1,] <- list(AveTuition2018[i,2], AveTuition2018[i,3], strtoi(AveTuition2018[i,6], base = 0L) + strtoi(AveTuition2018[i + 1,6], base = 0L))
  }
}
#Loops through AveTuition2018 to create a new data frame with removed columns that are unnecessary and adds a new column for the total cost of college  
```

Create three new college data tables for private, out-of-state, and in-state cost
```{r}
AveCostPrivate2018 <- filter(AveCost2018, Type %in% c("Private"))
#Filters for only private cost

AveCostIn2018 <- filter(AveCost2018, Type %in% c("Public In-State"))
#Filters for only in-state cost

AveCostOut2018 <- filter(AveCost2018, Type %in% c("Public Out-of-State"))
#Filters for only out-of-state cost

AveCostPrivate2018 <- AveCostPrivate2018[,-2]
#Removes type column since they are all the same type now

AveCostIn2018 <- AveCostIn2018[,-2]
#Removes type column since they are all the same type now

AveCostOut2018 <- AveCostOut2018[,-2]
#Removes type column since they are all the same type now

AveCostPrivate2018
AveCostIn2018
AveCostOut2018
```
Create plots comparing average annual income and cost of college for each state
```{r}
plot(AveIncome2018NoWyoming$Income, AveCostPrivate2018$Cost, main="Private Colleges", xlab="Average Annual Income (USD)", ylab="Private College Cost (USD)")
plot(AveIncome2018$Income, AveCostIn2018$Cost, main="In-State Colleges", xlab="Average Annual Income (USD)", ylab="In-State College Cost (USD)")
plot(AveIncome2018NoDC$Income, AveCostOut2018$Cost, main="Out-of-State Colleges", xlab="Average Annual Income (USD)", ylab="Out-of-State College Cost (USD)")
```
Create linear models for the above relationships
```{r}
print("Private Colleges")
private <- lm(AveCostPrivate2018$Cost~AveIncome2018NoWyoming$Income)
summary(private)
cor(AveIncome2018NoWyoming$Income, AveCostPrivate2018$Cost)

print("In-State Colleges")
inState <- lm(AveCostIn2018$Cost~AveIncome2018$Income)
summary(inState)
cor(AveIncome2018$Income, AveCostIn2018$Cost)

print("Out-of-State Colleges")
outState <- lm(AveCostOut2018$Cost~AveIncome2018NoDC$Income)
summary(outState)
cor(AveIncome2018NoDC$Income, AveCostOut2018$Cost)
```
Add regression line to plots comparing average annual income and cost of college for each state
```{r}
plot(AveIncome2018NoWyoming$Income, AveCostPrivate2018$Cost, main="Private Colleges", xlab="Average Annual Income (USD)", ylab="Private College Cost (USD)")
abline(private, col = "blue")
plot(AveIncome2018$Income, AveCostIn2018$Cost, main="In-State Colleges", xlab="Average Annual Income (USD)", ylab="In-State College Cost (USD)")
abline(inState, col = "blue")
plot(AveIncome2018NoDC$Income, AveCostOut2018$Cost, main="Out-of-State Colleges", xlab="Average Annual Income (USD)", ylab="Out-of-State College Cost (USD)")
abline(outState, col = "blue")
```

